jax-cosmo: End-to-End Differentiable and GPU Accelerated Cosmology Library

### 2020-08-21 paper.tex

Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
samples from high dimensional parameter spaces.  It was introduced in \citet{duane}
and developed extensively since.  See \citep{betancourt} for a full review; we describe
very basic features here.

HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
as the negative potential energy of a particle at each point $p$ in the space.  At each sample a trajectory are initialized
with a random velocity $q$, and then integrated using Hamilton's equations with a numerical integration:
\begin{align}
\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial {\cal P}}{\mathrm{d} q}\\
\frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
\end{align}
where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also
used to set the scale of the random initial velocities.

Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
some number of integration steps a final point in the trajectory is compared to the initial one,
and a Metropolis-Hastings acceptance criterion on the total energy $p^T M^{-1} p + {\cal}P(q)$ is applied.
If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
a relaxation to the integration accuracy.

That gradients $\frac{\partial {\cal P}}{\mathrm{d} q}$ can be estimated using finite differencing,
but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
calculating a Hessian, as in JAX-Cosmo, makes it feasible and efficient.

Below we show an example set of HMC constraints using JAX-Cosmo to constrain a simulated
Dark Energy Survey Year 1 (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
decomposition,
of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
edge of the parameter space, as recommended in \citet{afshar}.  We use a fixed number (25) of integration steps and 
a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
far from optimal.  We compare to the highly-optimized \emph{Cobaya} 
Metropolis-Hastings implementation \citep{cobaya1,cobaya2}, initializing it with the Fisher matrix as a proposal and
tuning thereafter.

The Hamiltonian sampler required about half the number of samples as the Metropolis to reach approximately the same
convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
on a CPU.  This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.  On next generation GPU high-performance systems, the advantage will be large.

>>>>

Hamiltonian Monte Carlo (HMC) is an MCMC-type method particularly suited to drawing
samples from high dimensional parameter spaces.  It was introduced in \citet{duane}
and developed extensively since.  See \citet{betancourt} for a full review; we describe
very basic features here.

HMC samples spaces by generating particle tracjectories through the space, using the log-posterior
as the negative potential energy of a particle at each point $q$ in the space.  At each sample a trajectory are initialized
with a random momentum $p$, and then integrated using Hamilton's equations with a numerical integration:
\begin{align}
\frac{\mathrm{d}p}{\mathrm{d}t} &= - \frac{\partial V}{\mathrm{d} q} = \frac{\partial \log{\cal P}}{\mathrm{d} q}\\
\frac{\mathrm{d}q}{\mathrm{d}t} &= + \frac{\partial U}{\mathrm{d} p} = M^{-1} p
\end{align}
where $M$ is a mass matrix which should be set to approximate the covariance of the posterior. This is also
used to set the scale of the random initial velocities.

Formally, the set of $n_\mathrm{dim}$ moment are treated as new parameters, and after 
some number of integration steps a final point in the trajectory is compared to the initial one,
and a Metropolis-Hastings acceptance criterion on the total energy $H = p^T M^{-1} p + \log{\cal P}(q)$ is applied.
If the trajectory is perfectly simulation then this acceptance is unity, since energy is conserved; applying it allows
a relaxation to the integration accuracy.

That gradients $\partial \log{\cal P} / \mathrm{d} q$ can be estimated using finite differencing,
but this typically requires at least $4 n_{\mathrm{dim}} + 1$ posterior evaluations per point, greatly slowing it
in high dimension, and as with the Fisher forecasting is highly prone to numerical error. Automatically
calculating a Hessian, as in JAX-Cosmo, makes it feasible and efficient.

Below we show an example set of HMC constraints using JAX-Cosmo to constrain a simulated
Dark Energy Survey Year One (DES-Y1) lensing and clustering-like (3x2pt) likelihood, using 21 parameters\footnote{We omit the 
five clustering sample photometric redshift parameters}.  We re-parametrize the space to unit mass using the Cholesky 
decomposition
of the covariance as estimated with an initial Fisher matrix; this lets us reflect the trajectories when they reach the
edge of the parameter space, as recommended in \citet{afshar}.  We use a fixed number (25) of integration steps and 
a fixed integration period (0.02), this was chosen heuristically to give a reasonable acceptance rate, but is likely to be
far from optimal.  We compare to the highly-optimized {\sc Cobaya} 
Metropolis-Hastings implementation \citep{cobaya2,cobaya1}, initializing it with the Fisher matrix as a proposal and
tuning thereafter.  We ran two processes for each sampler, with four CPU threads per process.

The Hamiltonian sampler required around half the number of samples as the Metropolis (70,000 vs 140,000) to reach approximately the same
convergence, as tested by the Gelman-Rubin test \citep{gelman}, though each sample was itself a few times slower
on a CPU.  This proof-of-principle example can certainly be hugely improved using various extensions to HMC such as No U-Turn Sampling \citep{nuts} or sampling from the entire trajectories \cite{betancourt}.  On next generation GPU high-performance systems, the advantage will be large.
